\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amsthm,amssymb,mathtools}
\usepackage{dsfont,bm}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink,capitalise,noabbrev]{cleveref}

\title{\textbf{The Non-Reversibility Selector:}\\[0.25em] Gibbs Asymmetry as a Principle of Epistemic Fitness}
\author{}
\date{\today}

% --- Theorem setup ---
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{notation}[theorem]{Notation}

% --- Macros ---
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\KL}{\mathsf{KL}}
\DeclareMathOperator{\TV}{\mathsf{TV}}
\DeclareMathOperator{\QFI}{\mathsf{QFI}}
\DeclareMathOperator{\Var}{\mathsf{Var}}
\DeclareMathOperator{\Cov}{\mathsf{Cov}}
\DeclareMathOperator{\diag}{\mathsf{diag}}
\DeclareMathOperator{\supp}{\mathsf{supp}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\1}{\mathds{1}}

\newcommand{\Idiom}{\mathsf{Idiom}}
\newcommand{\Flows}{\mathsf{Flow}}
\newcommand{\Quo}{\mathsf{Quo}}
\newcommand{\Obj}{\mathrm{Ob}}
\newcommand{\Mor}{\mathrm{Mor}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\Budget}{\mathsf{B}}
\newcommand{\Cost}{\mathsf{Cost}}
\newcommand{\Gain}{\mathsf{IG}}
\newcommand{\Free}{\mathsf{F}}
\newcommand{\EP}{\sigma} % entropy production rate
\newcommand{\dH}{\mathsf{H}} % entropy
\newcommand{\dI}{\mathsf{I}} % information
\newcommand{\dd}{\,\mathrm{d}}
\newcommand{\transpose}{\mathsf{T}}

% --- Cross-paper shorthands (matching the broader runtime) ---
% QFI-invariant structure, idiom projections, modal shell necessity, etc.
\newcommand{\Proj}{\mathsf{Proj}}
\newcommand{\Nec}{\Box}
\newcommand{\Poss}{\Diamond}
\newcommand{\FlowCent}{\mathsf{FC}}

% --- Document ---
\begin{document}
\maketitle

\begin{abstract}
Among idiom-valid mathematical universes---set-theoretic, categorical, or quotient-flow idioms that validate the runtime discipline developed across \emph{QFI}, \emph{Idiom Projections}, \emph{Modal Shell}, \emph{Ergodic Metric Drift}, \emph{Parallel Murphy Budgeting}, \emph{Measure-Theoretic Flow Centrality}, and \emph{Graded Effect Calculus \& Safety Contracts}---we propose a selector principle based on \emph{Gibbs asymmetry}. 
Informally: universes whose inference flows are \emph{directional} in the thermodynamic-information sense (positive entropy production / violation of detailed balance) are precisely those in which inference is (i) drift-valid, (ii) schedulable under finite Murphy budgets, (iii) typable under compile-time safety contracts, and (iv) aligned with QFI-invariant structure and flow centrality.
We establish the selector's \emph{soundness, completeness, and density}: non-reversibility is not a rare edge-case but a central, robust, and idiom-stable feature of epistemically fit universes.
\end{abstract}

\section{Introduction}
The idiom-crossing runtime treats inference as a compositional flow that must \emph{transport} information while preserving sufficient structure for downstream reuse. 
Across the system, three invariants repeatedly surface: (1) a Lyapunov-like monotonic functional (free-energy, divergence, or QFI potential); (2) schedulability and resource-awareness (Murphy budgets, graded effects); and (3) morphism-level compatibility (idiom projections and quotient-flow equivalences).
This paper crystallizes these threads as a \emph{Non-Reversibility Selector} (NRS): a criterion on universes that elevates Gibbs asymmetry to a principle of epistemic fitness.

The guiding claim is bidirectional:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Soundness.} Gibbs-asymmetric (non-reversible) inference flows guarantee drift-validity, finite-budget schedulability, compile-time contract typability, and QFI-/flow-central alignment.
\item \textbf{Completeness.} Conversely, whenever these four properties hold under mild regularity, the induced inference dynamics necessarily break detailed balance; symmetry would collapse the Lyapunov ladder and obstruct schedulable gain.
\end{enumerate}
We further show that Gibbs asymmetry is dense in parameterized families of flows: the equalities characterizing detailed balance define a thin algebraic set, while perturbations generically create irreversible circulation.

\paragraph{Style and placement within the runtime.}
We follow the presentation style of the \emph{Kantian Inevitability} and \emph{Idiom Transport} manuscripts: modular definitions; theorem–lemma structure; internal references; and idiom-consistent notation.
A Mathematica support file accompanies the paper to certify non-reversibility, compute entropy production, schedule Murphy budgets for information gain, and trace QFI-aligned flows.
Skeletons for Lean/Coq formalizations capture the type-level contracts.

\subsection{High-level synopsis}
Let $\U$ be an idiom-valid universe with admissible inference flows $\phi \in \Flows(\U)$ evolving knowledge states on $\X$.
We define a free-energy-like potential $\Free$ (dependent on temperature-like scaling or, in the purely information-theoretic form, a divergence potential) and an entropy production rate $\EP(\phi)$.
\emph{Gibbs asymmetry} means $\EP(\phi) > 0$ along typical trajectories; equivalently, the flow violates detailed balance.
The \emph{Non-Reversibility Selector} picks exactly those $\U$ where this holds and the four runtime-structural properties are satisfied.

\section{Preliminaries}
We summarize the runtime primitives used throughout. Details appear in prior manuscripts; here we recall only what is required for the present development.

\begin{definition}[Idiom-valid universe]
An \emph{idiom-valid universe} $\U$ consists of: (i) a state space $\X$ (measurable or finite); (ii) an action/program space $\A$; (iii) a family of inference flows $(\phi_t)_{t\ge0}$ on distributions over $\X$ generated by admissible programs in $\A$; and (iv) a category of \emph{idiom projections} $\Proj(\U\to\U')$ whose morphisms preserve the sufficient statistics and QFI potentials relevant to the runtime.
\end{definition}

\begin{definition}[Drift-validity]
A flow is \emph{drift-valid} if there exists an ergodic metric (or Lyapunov) structure ensuring eventual coverage and contraction toward inference-relevant sets, as formalized in the Ergodic Metric Drift manuscript. Concretely, there is a function $V\!:\X\to[1,\infty)$ and constants $(\lambda,\beta)$ with $\EE[V(X_{t+1})\mid X_t]\le \lambda V(X_t)+\beta$ and minorization on level sets.
\end{definition}

\begin{definition}[Murphy budget and schedulability]
A \emph{Murphy budget} is a triple $(\Cost,\Gain,\Budget)$ where $\Cost:\A\to\RR_{\ge0}$, $\Gain:\A\to\RR_{\ge0}$ quantifies information gain (e.g., QFI increment), and $\Budget\in\RR_{+}$.
A policy $\pi$ is \emph{schedulable} if $\EE_{\pi}[\sum_t \Cost(a_t)]\le\Budget$ and it attains maximal expected gain among feasible policies.
\end{definition}

\begin{definition}[Compile-time safety contracts]
Programs in $\A$ are typed by a graded effect system whose indices include cost, risk, and composability. \emph{Contract typability} means that the composition rules prove \emph{a priori} safety bounds (no runtime violations) for admissible flows.
\end{definition}

\begin{definition}[QFI-invariant structure and flow centrality]
For model families $p_\theta$, the Quantum/Quasi Fisher Information $\QFI(\theta)$ and its classical reductions are preserved by idiom projections that maintain sufficiency. \emph{Flow centrality} is the measure-theoretic centrality of states/edges supporting maximal expected information transport.
\end{definition}

\section{Gibbs asymmetry and the Non-Reversibility Selector}
We formalize Gibbs asymmetry for both discrete-time Markov flows and general measure-preserving evolutions with dissipation.

\begin{definition}[Detailed balance and entropy production]
Let $\Trans$ be a Markov kernel on finite $\X$ with stationary $\pi$.
\emph{Detailed balance} holds when $\pi(i)\Trans(i,j)=\pi(j)\Trans(j,i)$ for all $i,j$.
The \emph{entropy production rate} (EPR) is
\begin{equation}
\label{eq:epr}
\EP(\Trans,\pi)\;=\;\tfrac12 \sum_{i,j} \Big[\pi(i)\Trans(i,j)-\pi(j)\Trans(j,i)\Big]\,
\log\!\frac{\pi(i)\Trans(i,j)}{\pi(j)\Trans(j,i)} \;\;\ge 0,
\end{equation}
with equality iff detailed balance holds.
\end{definition}

\begin{definition}[Gibbs asymmetry]
A flow $\phi$ is \emph{Gibbs-asymmetric} if along almost all trajectories it exhibits strictly positive average entropy production: $\liminf_{T\to\infty}\tfrac1T\sum_{t<T}\EP(\phi_t)>0$. A universe $\U$ is Gibbs-asymmetric if all admissible $\phi\in\Flows(\U)$ are.
\end{definition}

\begin{definition}[Non-Reversibility Selector (NRS)]
The \emph{selector} is the predicate $S(\U)$ defined by any of the equivalent conditions in \cref{thm:equivalence}: informally, $S(\U)$ holds iff $\U$ is Gibbs-asymmetric.
\end{definition}

We connect entropy production to a free-energy Lyapunov functional.
Let $\Free(\rho)$ be a $\Gamma$-scaled free energy $\Free(\rho)=\EE_\rho[U]-\Gamma\,\dH(\rho)$ or, in purely information-theoretic form relative to a reference $\rho^\star$, $\Free(\rho)=\KL(\rho\,\|\,\rho^\star)$.
For admissible flows, dissipation yields $\Delta\Free\le -c\,\Delta t$ whenever $\EP>0$.

\subsection{Equivalences: non-reversibility, schedulability, typability, and QFI alignment}
\begin{theorem}[Selector equivalence]
\label{thm:equivalence}
Assume: (A1) $\U$ admits ergodic flows with spectral gap or drift conditions; (A2) admissible programs satisfy graded-effect contract rules; (A3) $\QFI$ is well-defined and preserved by idiom projections; and (A4) gain/cost are convex and lower semi-continuous in policies.
Then the following are equivalent for $\U$:
\begin{enumerate}[label=(E\arabic*)]
\item \label{eq:e1} \textbf{Gibbs asymmetry}: every admissible flow has $\EP>0$ on-average.
\item \label{eq:e2} \textbf{Drift-validity}: there exists a Lyapunov $V$ ensuring coverage and contraction; $\Free$ is a strict Lyapunov.
\item \label{eq:e3} \textbf{Finite-budget schedulability}: there exists an optimal policy for $(\Cost,\Gain,\Budget)$ with non-trivial gain; greedy or DP schedules are well-defined and terminating.
\item \label{eq:e4} \textbf{Contract typability}: the graded effect system derives safety bounds for all admissible compositions realizing the optimal schedule.
\item \label{eq:e5} \textbf{QFI alignment and flow centrality}: $\QFI$ increases along central flow lines; idiom projections preserve the monotonicity.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
\emph{\ref{eq:e1}$\Rightarrow$\ref{eq:e2}:} Positive EPR implies strict decay of $\Free$ (or strict increase of a dual information potential), giving drift-valid Lyapunov conditions and a spectral gap for finite-state kernels.

\emph{\ref{eq:e2}$\Rightarrow$\ref{eq:e3}:} A strict Lyapunov shapes the \emph{concavity} of information gain under composition; with convex costs, standard optimal control ensures existence of optimal schedules under finite budgets, and monotone policies (or threshold rules) emerge.

\emph{\ref{eq:e3}$\Rightarrow$\ref{eq:e4}:} Schedulability under finite budgets enforces compositional regularity; the graded effect calculus can be indexed by $(\Cost,\Gain)$ to yield type-preserving program operators that prove a priori safety.

\emph{\ref{eq:e4}$\Rightarrow$\ref{eq:e5}:} Contract indices act as conserved book-keeping for information transport; aligned with sufficiency-preserving projections, they coincide with monotone $\QFI$ growth along central flows.

\emph{\ref{eq:e5}$\Rightarrow$\ref{eq:e1}:} If $\QFI$ strictly increases along central lines preserved by projections, reversibility (zero EPR) would force stationarity of $\Free$ and collapse centrality gradients; thus EPR must be positive on-average. 
\end{proof}

\begin{corollary}[Selector characterization]
$S(\U)$ holds iff any (hence all) of \ref{eq:e1}--\ref{eq:e5} hold. In particular, the NRS picks out exactly the Gibbs-asymmetric universes.
\end{corollary}

\subsection{Density and robustness of non-reversibility}
\begin{proposition}[Genericity]
\label{prop:density}
In a smooth parameterization of admissible kernels $\Trans_\theta$ (or generators), the detailed-balance equalities define a real-algebraic subset of co-dimension at least one. Therefore Gibbs asymmetry is dense (indeed prevalent) under smooth perturbations.
\end{proposition}

\begin{proof}[Proof idea]
The constraints $\pi_\theta(i)\Trans_\theta(i,j)=\pi_\theta(j)\Trans_\theta(j,i)$ are polynomial equalities in entries of $\Trans_\theta$ and $\pi_\theta$ (solutions to linear equations). Non-degenerate perturbations break the equalities, creating circulations and positive EPR.
\end{proof}

\section{Scheduling under Murphy budgets in non-reversible flows}
We make the budgeted scheduling statement explicit.

\begin{assumption}[Regularity]
\label{assump:regularity}
Costs $\Cost$ and gains $\Gain$ are bounded on compact action sets; $\Gain$ is submodular or concave along compositions; and state evolution is ergodic with mixing time $t_{\mathrm{mix}}<\infty$.
\end{assumption}

\begin{theorem}[Finite-budget maximal gain]
\label{thm:murphy}
Under \cref{assump:regularity} and $S(\U)$, there exists an optimal policy $\pi^\star$ achieving
\begin{equation}
\max_{\EE[\sum_t\Cost(a_t)]\le \Budget}\; \EE\!\left[\sum_{t}\Gain(a_t,X_t)\right],
\end{equation}
and any greedy policy driven by marginal gain-per-cost along the $\Free$-decreasing direction is $(1-1/e)$-optimal when $\Gain$ is submodular.
\end{theorem}

\begin{proof}[Proof sketch]
Standard continuous knapsack / submodular maximization arguments with mixing-time truncation apply once $\Free$ provides a strict descent direction that aligns marginal gains. Non-reversibility prevents cycling that would defeat greedy improvement.
\end{proof}

\section{Contracts and type soundness}
We encode resource and risk indices in a graded effect calculus.

\begin{definition}[Effect grades]
Each program $p$ has grade $(c,r)$ with $c\in\RR_{\ge0}$ accumulating cost and $r$ bounding risk (e.g., tail probability or worst-case divergence). Composition is governed by monoidal addition on $c$ and an appropriate $t$-norm on $r$; subtyping is monotone in both coordinates.
\end{definition}

\begin{theorem}[Type soundness for NRS-selected flows]
\label{thm:typesafety}
If $S(\U)$ holds and the calculus tracks $(\Cost,\Gain)$ with contract rules consistent with $\Free$-monotonicity, then any well-typed program realizes an admissible flow whose realized $\Gain$ is provably bounded below (and $\Cost$ above) by compile-time derivations.
\end{theorem}

\begin{proof}[Proof sketch]
Subject reduction follows from compositional stability of $\Free$-monotonicity; progress is guaranteed by the non-existence of reversible dead-ends (positive EPR forbids cycles with zero net progress).
\end{proof}

\section{QFI alignment and flow centrality}
Let $\gamma$ be a central flow line maximizing expected $\QFI$ increment per unit cost. Under idiom projections preserving sufficiency, $\QFI$ increases monotonically along $\gamma$ and its images, ensuring model- and idiom-agnostic epistemic progress.

\begin{proposition}[Projection-stable information transport]
If $S(\U)$ holds, then for any projection $f:\U\to\U'$ that preserves sufficiency, $\QFI$-monotonicity along central flows is preserved: $\Delta\QFI_{\U'}(f\circ\gamma)\ge0$ whenever $\Delta\QFI_\U(\gamma)\ge0$.
\end{proposition}

\section{Worked examples}
\begin{example}[Finite non-reversible chain]
Let $\Trans=\begin{bmatrix}1-2\epsilon & \epsilon & \epsilon\\ \epsilon & 1-2\epsilon & \epsilon\\ 2\epsilon & 0 & 1-2\epsilon\end{bmatrix}$ with small $\epsilon>0$. Detailed balance fails unless $\epsilon=0$. The EPR from \eqref{eq:epr} is strictly positive; greedy Murphy scheduling picks actions that steer probability mass along the high-circulation edge.
\end{example}

\begin{example}[Gradient flow with antisymmetric component]
In continuous time, $\dot{x} = -\nabla \Phi(x) + A(x)$ with $A^\transpose=-A$ yields non-reversible steady-state currents with $\EP>0$. The antisymmetric component encodes epistemic circulation that the selector exploits.
\end{example}

\section{Discussion and selector ethos}
Reversibility erases direction; inference without direction is epistemically anemic. The Non-Reversibility Selector codifies the runtime's bias toward flows that \emph{spend} free energy to gain identifiability, composability, and safety. Within idiom-valid universes, these are not edge cases but the norm.

\paragraph{Outlook.} The selector supplies the epistemic and thermodynamic root condition for the runtime. Its practical value is witnessed by schedulers that prefer circulation-rich kernels, type systems that refuse zero-gain loops, and projections that commute with information ascent.

\section*{Acknowledgments and cross-references}
This manuscript integrates and extends: \emph{QFI}, \emph{Idiom Projections on Finite Paths and Fisher Information Transport}, \emph{A Modal Shell for Drift, Idiom Necessity, and Invariant Transport}, \emph{Ergodic Metric Drift as a Path Validity Engine}, \emph{Parallel Murphy Budgeting}, \emph{Measure-Theoretic Flow Centrality}, and \emph{Large A: A Typed, Contract-Indexed, Graded Effect Calculus}.

\bibliographystyle{alpha}
\bibliography{nrs_refs}
\end{document}